---
title: "San Juan de Fuca"
author: "Luis Ángel Rodríguez García"
date: "25-05-2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(echo = TRUE, root.dir = rprojroot::find_rstudio_root_file())
library(rprojroot)
library(here)
library(knitr)
library(rgl)
rgl::setupKnitr()
source(here('R', 'psc_sne.R'))
source(here('R', 'psc_sne.R'))
source(here('R', "perplexity.R"))
source(here('R', "low_dimension.R"))
source(here('R', "high_dimension.R"))
source(here('R', "prob_distribution.R"))
source(here('R', "utils.R"))
library(ncdf4)
library(lubridate)
library(circular)
library(dplyr)
library(abind)
library(tidyr)
print(here())
```

```{r}
juanfuca_pkg <-  paste(here("data-raw", "strait-juan-fuca"), "jdf.rda", sep = "/")
load(juanfuca_pkg)
```

Let's calculate some rho values for a perplexity of 12, 20, 33 and 45.

```{r cache=TRUE}

rho_12 <- rho_optim_bst(sanjuanfuca, 12)
rho_20 <- rho_optim_bst(sanjuanfuca, 20)
rho_33 <- rho_optim_bst(sanjuanfuca, 33)
rho_45 <- rho_optim_bst(sanjuanfuca, 45)

```

Let's reduce the dimension using the polyspherical Cauchy SNE. First, optimized rho for a perplexity of 12 and $d=1$:

```{r cache=TRUE}
Y <- psc_sne(X=sanjuanfuca, d=1, 
             rho_psc_list = rho_12, 
             num_iteration=300)
```

```{r fig.asp=1}
Y_rad <- DirStats::to_rad(Y)
r <- 1
theta <<- Y_rad
plot(r*sin(theta),
     r*cos(theta),
     xlim=c(-max(r),max(r)),
     ylim=c(-max(r),max(r)))

polygon(max(r)*sin(seq(0,2*pi,length.out=100)),max(r)*cos(seq(0,2*pi,length.out=100)))
```

Now, optimized rho for a perplexity of 12 and $d=2$:

```{r cache=TRUE}
Y <- psc_sne(X=sanjuanfuca, d=2, 
             rho_psc_list = rho_12, 
             num_iteration=300)
```

```{r}
scatterplot3d::scatterplot3d(Y, xlim = c(-1, 1), ylim = c(-1, 1), 
                             zlim = c(-1, 1), color = rep(1, nrow(Y)))

rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
             radius = 1, type = "s", col = "lightblue", alpha = 0.25,
             lit = FALSE)
rgl::points3d(Y, col = rep(1, nrow(Y)))
```
Now, optimized rho for a perplexity of 20 and $d=1$:

```{r cache=TRUE}
Y <- psc_sne(X=sanjuanfuca, d=1, 
             rho_psc_list = rho_20, 
             num_iteration=300)
```

```{r fig.asp=1}
Y_rad <- DirStats::to_rad(Y)
r <- 1
theta <<- Y_rad
plot(r*sin(theta),
     r*cos(theta),
     xlim=c(-max(r),max(r)),
     ylim=c(-max(r),max(r)))

polygon(max(r)*sin(seq(0,2*pi,length.out=100)),max(r)*cos(seq(0,2*pi,length.out=100)))
```

Now, optimized rho for a perplexity of 12 and $d=2$:

```{r cache=TRUE}
Y <- psc_sne(X=sanjuanfuca, d=2, 
             rho_psc_list = rho_12, 
             num_iteration=300)
```

```{r}
scatterplot3d::scatterplot3d(Y, xlim = c(-1, 1), ylim = c(-1, 1), 
                             zlim = c(-1, 1), color = rep(1, nrow(Y)))

rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
             radius = 1, type = "s", col = "lightblue", alpha = 0.25,
             lit = FALSE)
rgl::points3d(Y, col = rep(1, nrow(Y)))
```

Now, optimized rho for a perplexity of 33 and $d=1$:

```{r cache=TRUE}
Y <- psc_sne(X=sanjuanfuca, d=1, 
             rho_psc_list = rho_33, 
             num_iteration=300)
```

```{r fig.asp=1}
Y_rad <- DirStats::to_rad(Y)
r <- 1
theta <<- Y_rad
plot(r*sin(theta),
     r*cos(theta),
     xlim=c(-max(r),max(r)),
     ylim=c(-max(r),max(r)))

polygon(max(r)*sin(seq(0,2*pi,length.out=100)),max(r)*cos(seq(0,2*pi,length.out=100)))
```

Now, optimized rho for a perplexity of 33 and $d=2$:

```{r cache=TRUE}
Y <- psc_sne(X=sanjuanfuca, d=2, 
             rho_psc_list = rho_33, 
             num_iteration=300)
```

```{r}
scatterplot3d::scatterplot3d(Y, xlim = c(-1, 1), ylim = c(-1, 1), 
                             zlim = c(-1, 1), color = rep(1, nrow(Y)))

rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
             radius = 1, type = "s", col = "lightblue", alpha = 0.25,
             lit = FALSE)
rgl::points3d(Y, col = rep(1, nrow(Y)))
```

Lastly, optimized rho for a perplexity of 45 and $d=1$:

```{r cache=TRUE}
Y <- psc_sne(X=sanjuanfuca, d=1, 
             rho_psc_list = rho_45, 
             num_iteration=300)
```

```{r fig.asp=1}
Y_rad <- DirStats::to_rad(Y)
r <- 1
theta <<- Y_rad
plot(r*sin(theta),
     r*cos(theta),
     xlim=c(-max(r),max(r)),
     ylim=c(-max(r),max(r)))

polygon(max(r)*sin(seq(0,2*pi,length.out=100)),max(r)*cos(seq(0,2*pi,length.out=100)))
```

Now, optimized rho for a perplexity of 33 and $d=2$:

```{r cache=TRUE}
Y <- psc_sne(X=sanjuanfuca, d=2, 
             rho_psc_list = rho_45, 
             num_iteration=300)
```

```{r}
scatterplot3d::scatterplot3d(Y, xlim = c(-1, 1), ylim = c(-1, 1), 
                             zlim = c(-1, 1), color = rep(1, nrow(Y)))

rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
             radius = 1, type = "s", col = "lightblue", alpha = 0.25,
             lit = FALSE)
rgl::points3d(Y, col = rep(1, nrow(Y)))
```
