---
title: "San Juan de Fuca"
author: "Luis Ángel Rodríguez García"
date: "25-05-2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(echo = TRUE, root.dir = rprojroot::find_rstudio_root_file())
library(rprojroot)
library(here)
library(knitr)
library(rgl)
library(ncdf4)
library(lubridate)
library(circular)
library(dplyr)
library(abind)
library(tidyr)
library(pscsne)
library(webshot2)
library(ggquiver)
library(ggmap)

knitr::opts_chunk$set(echo = TRUE)
rgl::setupKnitr()
```

Transform the original data set into an array of dimension `c(n, d+1, r)`, where `n` is the time instant, `r` is the number of (latitude and longitude) coordinates and `d+1` are the Cartesian angles which in this cases is equal to 2. 

```{r}
# Get long format data from strait of Juan de Fuca
juanfuca_pkg <-  paste(here("data-raw", "strait-juan-fuca"), "jdf.rda", sep = "/")
load(juanfuca_pkg)

# Obtain other data frame
# Transform into coordinates (value formed by lat and lon separated by a comma)
# Remove speed from the dataset
jdf_coor <- jdf_long_fmt %>% 
  unite(col = "coordinates", lat, lon, sep = ",", remove = FALSE) %>% 
  dplyr::select(-"speed") 
  
# Transform the data frame into a 3-dimensional array
jdf_coor <- abind(split(jdf_coor, jdf_coor$coordinates), along = 3)


# Sum the number of NA's by each row
to_remove <- colSums(t(sapply(1:dim(jdf_coor)[3], 
                              function(k) is.na(jdf_coor[ , 5, k]))))
# Indexes of the rows without NA's
index_to_not_remove <- to_remove == 0

# Select only those rows without NA's
jdf_coor <- jdf_coor[index_to_not_remove , , ]

```

```{r eval = FALSE}
# Calculate the Cartesian angles
jdf_pscsne <- sapply(1:dim(jdf_coor)[3], 
                     function(k) DirStats::to_cir(as.numeric(jdf_coor[ , 3, k])), 
                     simplify = 'array')

# Get the rho values for a specific perplexity
perplexity <- 30
rho_30 <- rho_optim_bst(jdf_pscsne, perplexity, 7)
# Time difference of 1.590534 days

# Save dataset in long format
save(
  list = c("jdf_coor", "index_to_not_remove", "jdf_pscsne", "perplexity", "rho_30"),
  file = paste(here("data-raw", "strait-juan-fuca"), "pscsne.rda", sep = "/")
)

```

Let's load the data objects generated in the previous chunk.

```{r}
# Loading the psc-sne input object
load(paste(here("data-raw", "strait-juan-fuca"), "pscsne.rda", sep = "/"))
```

Now we are ready to reduce the dimension using the polyspherical Cauchy SNE. First, optimized rho is based on a perplexity of 30, $d=1$ and initializing the resultant reduced data with optimized evenly space points:

```{r cache=TRUE}
Y_1 <- psc_sne(X = jdf_pscsne, d = 1, rho_psc_list = rho_30, num_iteration = 500, 
             init = "equispaced")
```

```{r fig.asp=1}
Y_rad <- DirStats::to_rad(Y_1)
r <- 1
theta <<- Y_rad
plot(r * sin(theta), r * cos(theta), xlim = c(-max(r), max(r)), 
     ylim = c(-max(r), max(r)))

polygon(max(r) * sin(seq(0,2 * pi, length.out = 100)),
        max(r) * cos(seq(0,2 * pi, length.out = 100)))
```

We can not get so many insights from the previous plot, all the points are around the circumference. Let's see reducing the data to a sphere.

Let's now reduced the dimension onto the sphere, $d=2$, of radius 1, with an optimized rho based on a perplexity of 30 and initialize the resultant reduced dimension data with optimized evenly space points:

Let's do the same but initializing the resultant reduced dimension data with optimized evenly space points:

```{r cache=TRUE}
Y <- psc_sne(X = jdf_pscsne, d = 2, rho_psc_list = rho_30, num_iteration = 500, 
             init = "equispaced")
```

And plot the results in an sphere:

```{r}
scatterplot3d::scatterplot3d(Y, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1), 
                             color = rep(1, nrow(Y)))
```

Onto the 3d sphere. Here we can see a well-defined belt which separates observations.

```{r, fig.height=5, fig.width=5, fig.align='center', rgl=TRUE, dev='png'}
rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
             radius = 1, type = "s", col = "lightblue", alpha = 0.8, lit = FALSE)
rgl::points3d(Y, col = rep(1, nrow(Y)))
rgl.viewpoint(theta = -90, phi = -90, zoom = .9)
```


Nevertheless, as you can see in the next plot, there is some continuity so the belt is not completely around the sphere. That could be informing the difference between "East" and "West" flows and also on the existence of other intermediate regimes.

```{r, fig.height=5, fig.width=5, fig.align='center', rgl=TRUE, dev='png'}
rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
             radius = 1, type = "s", col = "lightblue", alpha = 0.8, lit = FALSE)
rgl::points3d(Y, col = rep(1, nrow(Y)))
rgl.viewpoint(theta = 180, phi = 90, zoom = .9)
```

Let's see what are the rows located in both sides of the belt, to figure out why they are well separated.

```{r, eval = FALSE}
rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
            radius = 1, type = "s", col = "lightblue", alpha = 1, lit = FALSE)
rgl::text3d(Y, text = 1:nrow(Y), cex = 0.8)
rgl.viewpoint(theta = -90, phi = -90, zoom = .3)
```

As it is shown in the previous plot (which is a zoom-in in the zone around the belt), there are two groups:

- The one group which is above the belt, with the following row ids: *2157*, *2215*, *2205*, *2217*, *3168*, *2166* and *3004*.
- The other group is below the well-separated area, with the following row ids: *988*, *288*, *1848*, *467*, *188*, *671* and *173*.

```{r}
up_group <- c(2157, 2215, 2205, 2217, 3168, 2166, 3004)
below_group <- c(988, 288, 1848, 467, 188, 671, 173)

total <- c(up_group, below_group)
colors <- c(rep(2, length(up_group)), rep(3, length(below_group)))

data_groups <- jdf_pscsne[ total, , ]

par(mfrow = c(3, 5))
for(k in 1:dim(jdf_pscsne)[3]) {
  
  Y_rad <- DirStats::to_rad(data_groups[ , , k])
  r <- 1
  theta <<- Y_rad
  plot(r*sin(theta), r*cos(theta), col = colors, xlim = c(-max(r), max(r)),
       ylim = c(-max(r), max(r)))
  polygon(max(r) * sin(seq(0,2 * pi, length.out = 100)), 
          max(r) * cos(seq(0,2 * pi,length.out = 100)))
  
}

```

Let's see if the package `movMF` can detect automatically the two mixtures:

```{r}
library(movMF)
Y_rad <- DirStats::to_rad(Y)
fitted <- movMF(x = Y_rad, k = 2, nruns = 10)
colors <- predict(fitted)

rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
             radius = 1, type = "s", col = "lightblue", alpha = 1, lit = FALSE)
rgl::points3d(Y, col = colors)
```

It didn't succed, then we can pick up two points that represent the north and the south poles and, later on, assign a group to the remainder points.


```{r}

north_point <- 3609
south_point <- 2088

cos_sim_sph <- drop(sphunif::Psi_mat(array(Y, dim = c(nrow(Y), ncol(Y), 1)), 
                                     scalar_prod = TRUE))
cos_sim_mat <- vec2matrix(cos_sim_sph, n = nrow(Y), diag_value = 1)

north_sounth_cos_sim <- cos_sim_mat[c(north_point, south_point), ]

# distance in the sphere of radius 1 -> cos^{-1}(A*B)
colors <- apply(X = acos(north_sounth_cos_sim), MARGIN = 2, FUN = which.min)

rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
             radius = 1, type = "s", col = "lightblue", alpha = 1, lit = FALSE)
rgl::points3d(Y, col = colors)
```

Let's use this color to see if it could split the two groups in the first reduction dimension we have done, where $d = 1$:

```{r}
Y_rad <- DirStats::to_rad(Y_1)
r <- 1
theta <<- Y_rad
plot(r * sin(theta), r * cos(theta), col = colors, xlim = c(-max(r), max(r)), 
     ylim = c(-max(r), max(r)))

polygon(max(r) * sin(seq(0,2 * pi, length.out = 100)),
        max(r) * cos(seq(0,2 * pi, length.out = 100)))
```

Now that we have concluded that it is clear that our algorithm is clustering the data along the circumference, we can obtain the kde of the reduced dimension data:

```{r}
library(ks)
fhat <- ks::kde(x = Y_1)
plot(fhat)
```

Now, it is time to visualize how the current sea evolves for the diferent instants depending on the group assigned:

```{r cache = TRUE}
num_instants <- nrow(jdf_coor)
num_coordinates <- dim(jdf_coor)[3]

jdf_long_fmt_res <- NULL
for(k in 1:num_coordinates) {
  
  jdf_long_fmt_res <- rbind(jdf_long_fmt_res, jdf_coor[ , , k])
  
}


# Delete the column coordinates
jdf_long_fmt_res <- jdf_long_fmt_res[,-1]

jdf_long_fmt_res <- data.frame(jdf_long_fmt_res) %>% 
  mutate(
    time = as.POSIXct(time, origin = "UTC"),
    lat = as.numeric(lat),
    lon = as.numeric(lon),
    theta = as.numeric(theta)
  ) %>% 
  arrange(time)

hours <- 3
instant_seq <- seq(
    from = as.POSIXct("2020-06-01 00:00:00", tz = "UTC"), 
    to = as.POSIXct("2022-07-01 00:00:00", tz = "UTC"),
    by = paste(hours, "hours", sep = " ")
  )
instant_seq <- instant_seq[-length(instant_seq)]
instant_seq <- instant_seq[index_to_not_remove]

jdf_long_fmt_strtime <- jdf_long_fmt_res %>% 
  mutate(time = as.character(time))

jdf_by_time_res <- abind(split(jdf_long_fmt_strtime, jdf_long_fmt_strtime$time), along = 3)

unique_lat <- unique(jdf_long_fmt_res$lat)
unique_lon <- unique(jdf_long_fmt_res$lon)
lat_length <- length(unique_lat)
lon_length <- length(unique_lon)
lat_values <- rev(sort(unique_lat))
lon_values <- sort(unique_lon)

r <- num_instants
n <- num_coordinates

jdf_by_time_latlon <- array(NA, dim = c(lon_length, lat_length, dim(jdf_by_time_res)[3]))
colors_by_time_latlon <- array(NA, dim = c(lon_length, lat_length, dim(jdf_by_time_res)[3]))

for(k in seq_len(r)) {
  
  for(i in seq_len(n)) {
      
    val <- jdf_by_time_res[i, ,k]
    lat <- as.numeric(val["lat"])
    lon <- as.numeric(val["lon"])
    index_lat <- which(round(lat_values, digits = 5) == lat)
    index_lon <- which(round(lon_values, digits = 4) == lon)
    theta <- as.numeric(val["theta"])
    
    jdf_by_time_latlon[index_lon, index_lat, k] = theta
    colors_by_time_latlon[index_lon, index_lat, k] = ifelse(colors[k] == 1, 3, 4)
    
  }
  
}

colnames(jdf_by_time_latlon) <- lat_values
rownames(jdf_by_time_latlon) <- lon_values

plot_vfield <- function(k) {

  loc = "Juan de Fuca"
  tim_i <- format(instant_seq[k], "%Y-%m-%d %H")
  direction <- ifelse(colors[k] == 1, "S/E", "N/W")
  title <- paste(loc, tim_i, "\n", direction)
  OceanView::quiver2D(cos(jdf_by_time_latlon[ , , k]), sin(jdf_by_time_latlon[ , , k]), 
                      x = lon_values,
                      y = lat_values,
                      col = colors_by_time_latlon[ , , k], clim = c(0, 1),
                      xlim = c(min(lon_values - 0.015), max(lon_values + 0.015)),
                      ylim = c(min(lat_values - 0.015), max(lat_values + 0.015)),
                      main = title, colkey = FALSE, xlab = "Longitude", 
                      ylab = "Latitude")
  points(expand.grid(lon_values, lat_values), pch = 16, cex = 0.5)

}

par(mfrow = c(3, 3))
for (k in seq_len(length(instant_seq))) plot_vfield(k)

```

```{r eval = FALSE}

# manipulate::manipulate(plot_vfield(k), k = manipulate::slider(1, r, step = 1))


animation::saveVideo(expr = {
  animation::ani.options(interval = 0.05, ani.res = 1080, ani.width = 1080, ani.height = 1080)
  times_seq <- 1:length(instant_seq)
  for (i in times_seq) {
    plot_vfield(i)
  }
}, video.name = "juandefuca.mp4", ffmpeg = "/opt/homebrew/opt/ffmpeg@2.8/bin/ffmpeg")
```


```{r eval = FALSE}
 
jdf_long_fmt_res_uv <- jdf_long_fmt_res %>% 
  mutate(
    u = cos(theta),
    v = sin(theta),
    time = as.character(time),
    color = rep(colors, each = dim(jdf_coor)[3])
  )

jdf_by_time_uv <- abind(split(jdf_long_fmt_res_uv, jdf_long_fmt_res_uv$time), along = 3)
jdf_by_time_uv <- jdf_by_time_uv[, -c(3, 4), ]

(map <<- get_map(c(left = min(lon_values) - 0.1, 
                    bottom = min(lat_values) - 0.1,
                    right = max(lon_values) + 0.1, 
                    top = max(lat_values) + 0.1)))

dput(map, file = "myMaps")

plot_ggvfield <- function(i) {
  data <- data.frame(jdf_by_time_uv[ , , i]) %>% 
    mutate(
      lon = as.numeric(lon),
      lat = as.numeric(lat),
      u = as.numeric(u),
      v = as.numeric(v),
      color = as.numeric(color)
    ) %>% 
    select(lon, lat, u, v, color)
  
  loc = "Juan de Fuca"
  tim_i <- format(instant_seq[i], "%Y-%m-%d %H")
  direction <- ifelse(colors[i] == 1, "S/E", "N/W")
  title <- paste(loc, tim_i, "\n", direction)
  
  ggobject <- NULL
  if(first(data$color) == 1) {
    
    ggmap(dget(file = "myMaps"), extent = "panel", geom = "blank", zoom = 0.9, 
          maptype = "toner-lite") + 
    geom_quiver(mapping = aes(u = u, v = v, color = 4), data = data, 
                center = TRUE, show.legend = FALSE) + ggtitle(title)
    
  } else {
    
    ggmap(dget(file = "myMaps"), extent = "panel", geom = "blank", zoom = 0.9, 
            maptype = "toner-lite") +
      geom_quiver(mapping = aes(u = u, v = v), data = data, 
                  center = TRUE, show.legend = FALSE) + ggtitle(title)
    
  }
  ggsave(paste(here("data-raw", "strait-juan-fuca", "img"), paste0("map", i, ".png"), sep = "/"))
}

times_seq <- 1:dim(jdf_by_time_uv)[3]
for( i in times_seq) {
  
  plot_ggvfield(i)
  
}

# ffmpeg -y -r 1/0.05 -i map%d.png juandefucamap.mp4
```
