---
title: "Applications PSC-SNE"
author: "Luis Ángel Rodríguez García"
date: "27-05-2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rprojroot)
library(here)
library(knitr)
library(rgl)
library(mvtnorm)
rgl::setupKnitr()
```

```{r include=FALSE}
knitr::opts_knit$set(echo = TRUE, root.dir = rprojroot::find_rstudio_root_file())
source(here('R', 'psc_sne.R'))
source(here('R', 'psc_sne.R'))
source(here('R', "perplexity.R"))
source(here('R', "low_dimension.R"))
source(here('R', "high_dimension.R"))
source(here('R', "prob_distribution.R"))
source(here('R', "utils.R"))
```

## Cases

### Case 1

Sample on the $(\mathbb{S}^1)^2$ where $p=1$ and $r=2$.

Data:

```{r cache=TRUE}
# Sample on the (S^1)^2
n <- 200
vmf11 <- rotasym::r_vMF(n = n, mu = drop(DirStats::to_cir(th = 0)), kappa = 10)
vmf12 <- rotasym::r_vMF(n = n, mu = drop(DirStats::to_cir(th = pi)), kappa = 10)
x1 <- sdetorus::toPiInt(cbind(DirStats::to_rad(vmf11),
                              DirStats::to_rad(vmf12)))

vmf21 <- rotasym::r_vMF(n = n, mu = drop(DirStats::to_cir(th = pi / 2)), kappa = 5)
vmf22 <- rotasym::r_vMF(n = n, mu = drop(DirStats::to_cir(th = 0)), kappa = 5)
x2 <- sdetorus::toPiInt(cbind(DirStats::to_rad(vmf21),DirStats::to_rad(vmf22)))

x <- rbind(x1, x2)

# Cartesian coordinates
n <- 400
x_array <- array(dim = c(n, 2, 2))
x_array[, , 1] <- DirStats::to_cir(x[, 1])
x_array[, , 2] <- DirStats::to_cir(x[, 2])

indexes <- sample(1:n)
x_array <- x_array[indexes,,]
colors <- rep(c(1, 2), each = n/2)[indexes]

plot(x, xlim = c(-pi, pi), ylim = c(-pi, pi), axes = FALSE,
     col = rep(c(1, 2), each = n/2), pch = 16, 
     xlab = "x", ylab = "y")
sdetorus::torusAxis()

```

Let's calculate the rho parameters based on a perplexity of 20:

```{r cache=TRUE}
# Time difference of 2.582784 mins
rho_first_20 <- rho_optim_bst(x_array, 20)
```

First, let's reduce to dimension $\mathbb{S}^1$ then $d=1$ (circumference):

```{r cache=TRUE, fig.asp=1}
Y <- psc_sne(X=x_array, d=1, rho_psc_list = rho_first_20, num_iteration=125, 
             colors=colors, visualize_prog = TRUE)
```

```{r fig.asp=1}
Y_rad <- DirStats::to_rad(Y)
r <- 1
theta <- Y_rad

plot(r*sin(theta),
     r*cos(theta),
     col=colors,
     xlim=c(-max(r),max(r)),
     ylim=c(-max(r),max(r)))

polygon(max(r)*sin(seq(0,2*pi,length.out=100)),max(r)*cos(seq(0,2*pi,length.out=100)))
```

Now we are going to reduce to dimension $\mathbb{S}^2$ then $d=2$ (sphere):

```{r cache=TRUE, fig.asp=1}
Y <- psc_sne(X=x_array, d=2, rho_psc_list = rho_first_20, num_iteration=125,
             colors=colors, visualize_prog = TRUE, eta=100, check=F)
```

```{r fig.asp=1}
scatterplot3d::scatterplot3d(Y, xlim = c(-1, 1), ylim = c(-1, 1), 
                             zlim = c(-1, 1), color = colors)
```

It's clearly clustered the two groups we had in the original generated data.

### Case 2

Sample on the $(\mathbb{S}^2)^2$ where $p=2$ and $r=2$.


```{r cache=TRUE}
n=200
d=2
r=2
samp1 <- rotasym::r_vMF(n = n, mu = drop(DirStats::to_sph(th = 0, ph = 0.5)),
                        kappa = 50)
samp2 <- rotasym::r_vMF(n = n, mu = drop(DirStats::to_sph(th = 2, ph = -1.5)),
                        kappa = 50)
samp3 <- rotasym::r_vMF(n = n, mu = drop(DirStats::to_sph(th = -1, ph = 0)),
                        kappa = 50)
x1 <- rbind(samp1, samp2, samp3)

scatterplot3d::scatterplot3d(x1,
                             xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
                             color = rep(1:3, each = n), main="Sphere 1")

rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
             radius = 1, type = "s", col = "lightblue", alpha = 0.25,
             lit = FALSE)
rgl::points3d(x1, col = rep(c(1,2,3), each=n))

samp4 <- rotasym::r_vMF(n = n, mu = drop(DirStats::to_sph(th = 0, ph = 0.55)),
                        kappa = 60)
samp5 <- rotasym::r_vMF(n = n, mu = drop(DirStats::to_sph(th = 2, ph = -1.41)),
                        kappa = 60)
samp6 <- rotasym::r_vMF(n = n, mu = drop(DirStats::to_sph(th = -1.1, ph = 0.05)),
                        kappa = 60)
x2 <- rbind(samp4, samp5, samp6)

scatterplot3d::scatterplot3d(x2,
                             xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
                             color = rep(1:3, each = n), main="Sphere 2")
rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
             radius = 1, type = "s", col = "lightblue", alpha = 0.25,
             lit = FALSE)
rgl::points3d(x2, col = rep(c(1,2,3), each=n))


x_2 <- array(dim = c(n*3, 3, 2))
x_2[, , 1] <- x1
x_2[, , 2] <- x2

n <- nrow(x_2)
indexes <- sample(1:n)
x_2 <- x_2[indexes,,]
colors <- rep(c(1, 2, 3), each = n / 3)[indexes]
```

Let's calculate the rho parameters based on a perplexity of 20:

```{r cache=TRUE}
rho_second_perp20 <- rho_optim_bst(x_2, 20)
```

First, let's reduce to dimension $\mathbb{S}^1$ then $d=1$ (circumference):

```{r cache=TRUE, fig.height=10, fig.width=10}
Y <- psc_sne(X=x_2, d=1, rho_psc_list = rho_second_perp20, num_iteration=200,
             colors=colors, visualize_prog = TRUE)
```

```{r}
Y_rad <- DirStats::to_rad(Y)
r <- 1
theta <<- Y_rad
plot(r*sin(theta),
     r*cos(theta),
     col=colors,
     xlim=c(-max(r),max(r)),
     ylim=c(-max(r),max(r)))

polygon(max(r)*sin(seq(0,2*pi,length.out=100)),max(r)*cos(seq(0,2*pi,length.out=100)))
```

Now we are going to reduce to dimension $\mathbb{S}^2$ then $d=2$ (sphere):

```{r cache=TRUE, fig.height=10, fig.width=10}
Y <- psc_sne(X=x_2, d=2, rho_psc_list = rho_second_perp20, num_iteration=200,
             colors=colors, visualize_prog = TRUE)
```


```{r fig.asp=1}
scatterplot3d::scatterplot3d(Y, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1), 
                             color = colors)
```

It's clearly clustered the three groups we had in the original generated data.


# Case 3

Sample on the $(\mathbb{S}^1)^3$ where $p=1$ and $r=3$.

```{r cache=TRUE}
# Sample on the (S^1)^2
n_ori <- 200
vmf11 <- rotasym::r_vMF(n = n_ori, mu = drop(DirStats::to_cir(th = 0)), kappa = 10)
vmf12 <- rotasym::r_vMF(n = n_ori, mu = drop(DirStats::to_cir(th = pi)), kappa = 10)
vmf13 <- rotasym::r_vMF(n = n_ori, mu = drop(DirStats::to_cir(th = pi/2)), kappa = 10)
x1 <- sdetorus::toPiInt(cbind(DirStats::to_rad(vmf11),
                              DirStats::to_rad(vmf12),
                              DirStats::to_rad(vmf13)))

vmf21 <- rotasym::r_vMF(n = n_ori, mu = drop(DirStats::to_cir(th = pi / 2)), kappa = 5)
vmf22 <- rotasym::r_vMF(n = n_ori, mu = drop(DirStats::to_cir(th = 0)), kappa = 5)
vmf23 <- rotasym::r_vMF(n = n_ori, mu = drop(DirStats::to_cir(th = pi)), kappa = 5)
x2 <- sdetorus::toPiInt(cbind(DirStats::to_rad(vmf21),DirStats::to_rad(vmf22), 
                              DirStats::to_rad(vmf23)))

vmf31 <- rotasym::r_vMF(n = n_ori, mu = drop(DirStats::to_cir(th = pi)), kappa = 30)
vmf32 <- rotasym::r_vMF(n = n_ori, mu = drop(DirStats::to_cir(th = 0)), kappa = 30)
vmf33 <- rotasym::r_vMF(n = n_ori, mu = drop(DirStats::to_cir(th = pi/2)), kappa = 30)
x3 <- sdetorus::toPiInt(cbind(DirStats::to_rad(vmf31),DirStats::to_rad(vmf32), 
                              DirStats::to_rad(vmf33)))

```

Now, let's do the same visualization with the data on the torus:

```{r fig.asp=1}
x <- rbind(x1, x2, x3)
n <- n_ori * 3
x_3 <- array(dim = c(n, 2, 3))
x_3[, , 1] <- DirStats::to_cir(x[, 1])
x_3[, , 2] <- DirStats::to_cir(x[, 2])
x_3[, , 3] <- DirStats::to_cir(x[, 3])

n <- nrow(x_3)
indexes <- sample(1:n)
x_3 <- x_3[indexes,,]
colors <- rep(c(1, 2, 3), each = n / 3)[indexes]


pairs(x[indexes,], col = colors)
scatterplot3d::scatterplot3d(x[indexes,], xlim = c(-pi, pi), ylim = c(-pi, pi), zlim = c(-pi, pi), 
                             color = colors)
```

Let's calculate the rho parameters based on a perplexity of 25:

```{r cache=TRUE}
rho_third_perp25 <- rho_optim_bst(x_3, 25)
```

First, let's reduce to dimension $\mathbb{S}^1$ then $d=1$ (circumference):

```{r cache=TRUE, fig.height=10, fig.width=10}
Y <- psc_sne(X=x_3, d=1, rho_psc_list = rho_third_perp25, num_iteration=200,
             colors=colors, visualize_prog = TRUE)
```

```{r fig.asp=1}
Y_rad <- DirStats::to_rad(Y)
r <- 1
theta <<- Y_rad
plot(r*sin(theta),
     r*cos(theta),
     col=colors,
     xlim=c(-max(r),max(r)),
     ylim=c(-max(r),max(r)))

polygon(max(r)*sin(seq(0,2*pi,length.out=100)),max(r)*cos(seq(0,2*pi,length.out=100)))
```

Now we are going to reduce to dimension $\mathbb{S}^2$ then $d=2$ (sphere):

```{r cache=TRUE, fig.height=10, fig.width=10}
Y <- psc_sne(X=x_3, d=2, rho_psc_list = rho_third_perp25, num_iteration=200,
             colors=colors, visualize_prog = TRUE)
```

```{r fig.asp=1}
scatterplot3d::scatterplot3d(Y, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1), 
                             color = colors)
```


# Case 4

```{r}
d <- 3
n <- 5e2
samp <- sdetorus::toPiInt(mvtnorm::rmvnorm(n = n, mean = rep(pi, d)))
pairs(samp)
scatterplot3d::scatterplot3d(samp, xlim = c(-pi, pi), ylim = c(-pi, pi), zlim = c(-pi, pi))

x_4 <- array(dim = c(n, 2, 3))
x_4[, , 1] <- DirStats::to_cir(samp[, 1])
x_4[, , 2] <- DirStats::to_cir(samp[, 2])
x_4[, , 3] <- DirStats::to_cir(samp[, 3])

```

Let's calculate the rho parameters based on a perplexity of 25:

```{r cache=TRUE}
rho_four_perp20 <- rho_optim_bst(x_4, 20)
```

First, let's reduce to dimension $\mathbb{S}^1$ then $d=1$ (circumference):

```{r cache=TRUE, fig.height=10, fig.width=10}
Y <- psc_sne(X=x_4, d=1, rho_psc_list = rho_four_perp20, num_iteration=200, 
             visualize_prog = TRUE)
```

```{r fig.asp=1}
Y_rad <- DirStats::to_rad(Y)
r <- 1
theta <<- Y_rad
plot(r*sin(theta),
     r*cos(theta),
     xlim=c(-max(r),max(r)),
     ylim=c(-max(r),max(r)))

polygon(max(r)*sin(seq(0,2*pi,length.out=100)),max(r)*cos(seq(0,2*pi,length.out=100)))
```

Now we are going to reduce to dimension $\mathbb{S}^2$ then $d=2$ (sphere):

```{r cache=TRUE, fig.height=10, fig.width=10}
Y <- psc_sne(X=x_4, d=2, rho_psc_list = rho_four_perp20, num_iteration=200,
             visualize_prog = TRUE)
```

```{r fig.asp=1}
scatterplot3d::scatterplot3d(Y, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1))
```


# Case 4

Playing with belts of points onto the sphere. First, let's generate the data

```{r}
rotate_matrix_z_axis <- function(alpha) {
  deg2rad <- function(deg) {(deg * pi) / (180)}
  rads <- deg2rad(alpha)
  matrix(c(cos(rads), -sin(rads), 0,
           sin(rads),  cos(rads), 0,
           0,                  0, 1),
         byrow = T, nrow = 3)
}

rotate_matrix_x_axis <- function(alpha) {
  deg2rad <- function(deg) {(deg * pi) / (180)}
  rads <- deg2rad(alpha)
  matrix(c(1,                 0 , 0,
           0, cos(rads), -sin(rads),
           0, sin(rads),  cos(rads)),
         byrow = T, nrow = 3)
}

rotate_matrix_y_axis <- function(alpha) {
  deg2rad <- function(deg) {(deg * pi) / (180)}
  rads <- deg2rad(alpha)
  matrix(c(cos(rads), 0, sin(rads),
           0,         1,         0,
          -sin(rads), 0, cos(rads)),
         byrow = T, nrow = 3)
}

p = 2
set.seed(2)
# A <- matrix(runif((p + 1)^2)*2-0.5, ncol = p + 1)
# sigma <- t(A) %*% A
sigma <- diag(c(10, 2, .5))
n <- 200
x_1_belt1 <- rotasym::r_ACG(n, sigma)
# TODO: sphunif::r_alt(n, alt = "SC", )
rotate_mat_90 <- rotate_matrix_z_axis(90)
x_2_belt1 <- x_1_belt1 %*% rotate_mat_90
rotate_mat_45 <- rotate_matrix_z_axis(45)
x_3_belt1 <- x_1_belt1 %*% rotate_mat_45
x_belt1 <- rbind(x_1_belt1, x_2_belt1, x_3_belt1)
scatterplot3d::scatterplot3d(x_belt1, xlim = c(-1, 1),
                             ylim = c(-1, 1), zlim = c(-1, 1),
                             color = rep(c(1,2,3), each=n))

rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
            radius = 1, type = "s", col = "lightblue",
            lit = FALSE)
rgl::points3d(x_belt1, col = rep(c(1,2,3), each=n))

x_1_belt2 <- rotasym::r_ACG(n, sigma)
x_2_belt2 <- x_1_belt2 %*% rotate_mat_90
x_3_belt2 <- x_1_belt1 %*% rotate_mat_45
x_belt2 <- rbind(x_1_belt2, x_2_belt2, x_3_belt2)
scatterplot3d::scatterplot3d(x_belt2, xlim = c(-1, 1),
                             ylim = c(-1, 1), zlim = c(-1, 1),
                             color = rep(c(1,2,3), each=n))

rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
            radius = 1, type = "s", col = "lightblue",
            lit = FALSE)
rgl::points3d(x_belt2, col = rep(c(1,2,3), each=n))

x_1_belt3 <- rotasym::r_ACG(n, sigma)
x_2_belt3 <- x_1_belt3 %*% rotate_mat_90
x_3_belt3 <- x_1_belt1 %*% rotate_mat_45
x_belt3 <- rbind(x_1_belt3, x_2_belt3, x_3_belt3)
scatterplot3d::scatterplot3d(x_belt3, xlim = c(-1, 1),
                             ylim = c(-1, 1), zlim = c(-1, 1),
                             color = rep(c(1,2,3), each=n))

rgl::plot3d(0, 0, 0, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
            radius = 1, type = "s", col = "lightblue",
            lit = FALSE)
rgl::points3d(x_belt3, col = rep(c(1,2,3), each=n))

n <- n * 3
x_5 <- array(dim = c(n, 3, 3))
x_5[, , 1] <- x_belt1
x_5[, , 2] <- x_belt2
x_5[, , 3] <- x_belt3

indexes <- sample(1:n)
x_5 <- x_5[indexes,,]
colors <- rep(c(1, 2, 3), each = n / 3)[indexes]
```

Let's calculate the rho parameters based on a perplexity of 25:

```{r}
rho_five_perp20 <- rho_optim_bst(x_5, 20)
```

First, let's reduce to dimension $\mathbb{S}^1$ then $d=1$ (circumference):

```{r fig.height=10, fig.width=10}
Y <- psc_sne(X=x_5, d=1, rho_psc_list = rho_five_perp20, num_iteration=200, 
             visualize_prog = TRUE, colors = colors)
```

```{r fig.asp=1}
Y_rad <- DirStats::to_rad(Y)
r <- 1
theta <<- Y_rad
plot(r*sin(theta),
     r*cos(theta),
     col=colors, 
     xlim=c(-max(r),max(r)),
     ylim=c(-max(r),max(r)))

polygon(max(r)*sin(seq(0,2*pi,length.out=100)),max(r)*cos(seq(0,2*pi,length.out=100)))
```

Now we are going to reduce to dimension $\mathbb{S}^2$ then $d=2$ (sphere):

```{r fig.height=10, fig.width=10}
Y <- psc_sne(X=x_5, d=2, rho_psc_list = rho_five_perp20, num_iteration=200,
             visualize_prog = TRUE, colors = colors)
```

```{r fig.asp=1}
scatterplot3d::scatterplot3d(Y, xlim = c(-1, 1), ylim = c(-1, 1), zlim = c(-1, 1),
                             color = colors)
```
